---
title: "EHEC, Introduction to R"
subtitle: "Orientation"
author: "Alex Hollingsworth, Grant McDermott, and Kelli Marquardt"
institute: "Indiana University | University of Oregon | Chicago FED"
date: "`r format(Sys.time(), '%d %B %Y')`" ## Or "Lecture no."
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css] 
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_light(base_color = "#23395b", link_color = "#F97B64")
```

```{r setup, include=FALSE, include=FALSE, cache=FALSE}
options(htmltools.dir.version = FALSE)
library(knitr)
opts_chunk$set(
  fig.align = "center",  
  fig.asp = 9/16,
  out.width = "95%",
  dpi= 300,
  cache=TRUE, 
  warning=FALSE, message=FALSE,
  echo=FALSE
  )
```


class: inverse, center, middle

# Prologue

---

## Requirements

☑ Installed [R](https://www.r-project.org/).

☑ Installed [RStudio](https://www.rstudio.com/products/rstudio/download/preview/).

--

Some optional (not necessary for today!) things that I recommend if you plan to use R going forward:

- **Windows:** Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/). 
- **Mac:**  Install [XQuartz](https://www.xquartz.org/) and the [R MacOS toolchain](https://github.com/rmacoslib/r-macos-rtools#how-do-i-use-the-installer).

--

These "extras" are mostly for when you want to build R packages directly from GitHub (e.g. development versions) that rely on compiled code (e.g. C++)

---

## Quick tour of RStudio

(Live session)

---

class: inverse, center, middle

# Why use / learn R?

---

## Pros

1. **Free.** Free as beer, free as in speech.

--

2. **Fast.** For many common operations, the R equivalents of (even optimised) Stata commands can be orders of magnitude faster. Examples: [reshaping](https://grantmcdermott.com/reshape-benchmarks/), [collapsing](https://twitter.com/grant_mcdermott/status/1402718339768545280), [HDFE regression](https://grantmcdermott.com/better-way-adjust-SEs/#benchmarking).

--

3. **Flexible.** A _huge_ array of packages and ancillary features. GIS, Machine learning, databases, version control (git), websites and apps, documents, etc. etc.

--

4. **~~Fobs~~Jobs.** Excellent outside options for a lot of econ graduates going into non-academic positions.

---

## Cons

1. **Unfamiliarity.** Learning a new language (can) suck. Go easy on yourself and R.

  - Quick tip: Stata thinks quite differently about programming and code than a lot of other languages. This doesn't make it bad! But it does mean that you might get stuck by trying to re-purpose your exact Stata workflow in R.

2. **Tyranny of information/choice.** R invariably has multiple ways to do exactly the same thing. Example: Assignment with `<-` and/or `=`.

  - Quick tip: Just pick whatever method/package looks easiest to you. At the same time, don't be scared to mix and match.
  
--

Meta thoughts: Pick whatever workflow works for you!

---

## Help

_Very_ friendly communities and forums online. SO, RLadies, Twitter, etc.

You can also look at the help documentation.

```r
?plot
example(plot)

library(ggplot2)
example(geom_contour)

demo("graphics", package = "graphics")
```

Most packages also provide _vignettes_ (mini use guides). [Example](https://lrberge.github.io/fixest/articles/fixest_walkthrough.html).

---

## Some other R differences/gotchas coming from Stata

- Multiple objects can exist in memory. This is the most common source of errors I see among newcomers.

- Log files aren't common (see point above), although can be used.

- You need to load packages into memory when you want to use them.

--

- Speaking of which, let's get our hands dirty with some examples...