---
title: "Introduction to R"
author: "Alex Hollingsworth, Grant McDermott, and Kelli Marquardt"
institute: "Indiana University | University of Oregon | Chicago FED"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      css: [kunoichi, metropolis, robot-fonts]
---
<style type="text/css">
.remark-slide-content {
    font-size: 25px;
    padding: 1em 4em 1em 4em;
}
</style>
# This document was made using R-Markdown.

- R-Markdown is great because you can combine text with R code and results. 

  - Text is formatted using markdown syntax, 
  - It's saved as an `.Rmd` file. 

- You can also just simply write `.R` code that is analogous to a `.do` file where there's a list of commands with comments. No need for this fancy `.Rmd` Markdown business. 

---
# This document was made using R-Markdown.

- It's pretty cool to integrate code with results right into the presentation. 

Input
```{r echo=FALSE, class.output="r", comment=""}
cat("```{r} \n sin(3) \n```")
```

--

Output
```{r}
sin(3)
```


---

# Learn more about R-Markdown

- If you are curious about markdown formatting, you can check out this handy guide. https://www.markdownguide.org/basic-syntax/

- To learn more about R-Markdown, this "cheat-sheet" https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf, is a good starting place. 

---
# Running code 

### Code Chunks
This is a code chunk that will evaluate the command  `sin(3)` 

```{r echo=FALSE, class.output="r", comment=""}
cat("```{r} \n sin(3) \n```")
```

The output of this will appear as follows:

```{r, echo = FALSE, eval = TRUE}
sin(3)
```

--

### Inline Code

You can also create calculations inline for example `` `r knitr::inline_expr("sin(3)")` `` will evaluate `sin(3)`. With output rendering like this `r sin(3)`.

---
# Let's switch over to an R-Markdown document
