---
title: "Introduction to R"
author: "Alex Hollingsworth, Grant McDermott, and Kelli Marquardt"
institute: "Indiana University | University of Oregon | Chicago FED"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      css: [kunoichi, metropolis, robot-fonts]
---
<style type="text/css">
.remark-slide-content {
    font-size: 25px;
    padding: 1em 4em 1em 4em;
}
</style>
# This document was made using R-Markdown.

- R-Markdown is great because you can combine text with R code and results. 

  - Text is formatted using markdown syntax, 
  - It's saved as an `.Rmd` file. 

- You can also just simply write `.R` code that is analogous to a `.do` file where there's a list of commands with comments. No need for this fancy `.Rmd` Markdown business. 

---
# This document was made using R-Markdown.

- It's pretty cool to integrate code with results right into the presentation. 

Input
```{r echo=FALSE, class.output="r", comment=""}
cat("```{r} \n sin(3) \n```")
```

--

Output
```{r}
sin(3)
```


---

# Learn more about R-Markdown

- If you are curious about markdown formatting, you can check out this handy guide. https://www.markdownguide.org/basic-syntax/

- To learn more about R-Markdown, this "cheat-sheet" https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf, is a good starting place. 

---
# Running code 

### Code Chunks
This is a code chunk that will evaluate the command  `sin(3)` 

```{r echo=FALSE, class.output="r", comment=""}
cat("```{r} \n sin(3) \n```")
```

The output of this will appear as follows:

```{r, echo = FALSE, eval = TRUE}
sin(3)
```

--

### Inline Code

You can also create calculations inline for example <code> r  sin(3) </code>
 will evaluate `sin(3)`. With output rendering like this `r sin(3)`.

---
# Loading packages
```{r}
library(ggplot2)
library(dplyr)
```

You can install packages using the `install.packages()` syntax, but see the section on `pacman` below before worrying about that. 

---
# Viewing data
```{r head-of-data} 
head(diamonds)
```

---
# Viewing data

We can get a list of column names fairly easily too

```{r list-of-column-names}
names(diamonds)
```

---
# Viewing data

```{r mean-price-by-color}
diamonds %>% group_by(color) %>% summarise(mean(price)) 
```

Here the pipe command `%>%` will take the output from the command on the left and "pipes" it as input to the command on the right. 

---
# Viewing data

```{r mean-price-by-color-cleaner}
diamonds %>% 
  group_by(color) %>% 
  summarise(mean(price)) 
```

1. You can just move onto a new line without an error. This is neat because it allows us to write cleaner code and you don't need a delimiter or the `///` you may be used to from stata. So we can rewrite the above like this. 
1. In more recent versions of R (4.1 and above), you don't need to have `dpylr` installed to pipe. You can do the same thing using `|>`. 

---
# Summary statistics table


<!--
# Let's switch over to an R-Markdown document

Personally, I love how xaringan looks. I like the simplicity. 

BUT I regularly run into minor road blocks and often am making presentations too close to the deadline, so end up switching to easier to use, but harder to update...Power point :)
-->