---
title: "Introduction to R"
author: "Alex Hollingsworth, Grant McDermott, and Kelli Marquardt"
institute: "Indiana University | University of Oregon | Chicago FED"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      css: [kunoichi, metropolis, robot-fonts]
---
# This document was made using R-Markdown.

- R-Markdown is great because you can combine text with R code and results. 

  - Text is formatted using markdown syntax, 
  - It's saved as an `.Rmd` file. 

- You can also just simply write `.R` code that is analogous to a `.do` file where there's a list of commands with comments. No need for this fancy `.Rmd` Markdown business. 

- But it's pretty cool to integrate code with results right in the presentation. 

```{r, eval = TRUE}
sin(3)
```

---

```{r}
sin(3)
```
- If you are curious about markdown formatting, you can check out this handy guide. https://www.markdownguide.org/basic-syntax/
- To learn more about R-Markdown, this "cheat-sheet" https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf, is a good starting place. 


---